---
title: "01-07-26 avian metapopulation model final"
output: html_document
date: "2026-07-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Modifications

This version has balanced population dynamics, standardised temperature using briere function and includes the bird breeding season. 1 infected bird is imported into each population in March of each year. Simulation runs over 5 years.

## Equations

$$
\frac{\mathrm{d}S_i}{\mathrm{d}t} = (\gamma N_i) - (\lambda S_i) - (\mu S_i)
$$ $$
\frac{\mathrm{d}E_i}{\mathrm{d}t} = (\lambda S_i)  - (\theta E_i) - (\mu E_i)
$$ $$
\frac{\mathrm{d}I_i}{\mathrm{d}t} = (\theta E_i) - (\mu I_i) - (\sigma I_i) + seed
$$ $$
\frac{\mathrm{d}R_i}{\mathrm{d}t} = (\sigma I_i) - (\mu R_i)
$$ $$
\lambda = briere(beta_{i,1} \frac {I_1}{N_1} + beta_{i,2} \frac {I_2}{N_2} + beta_{i,3} \frac {I_3}{N_3}+ beta_{i,4} \frac {I_4}{N_4}+ beta_{i,5} \frac {I_5}{N_5})
$$

## Matrix

$$
\begin{aligned}
beta &=
\begin{bmatrix}
3.0 & 0.01 & 0.01 \\
0.01 & 3.0 & 0.01 \\
0.01 & 0.01 & 3.0
\end{bmatrix}
\qquad
beta &=
\begin{bmatrix}
\beta_{11} & \beta_{12} & \beta_{13}\\
\beta_{21} & \beta_{22} & \beta_{23}\\
\beta_{31} & \beta_{32} & \beta_{33}
\end{bmatrix}
\end{aligned}
$$

low connectivity is represented by beta = 0.01 moderate connectivity is represented by beta = 0.05 high connectivity is represented by beta = 0.1

```{r matrix}
library(odin2)
library(dust2)
library(monty)
library(ggplot2)
library(tidyr)

# example beta = 0.1

beta <- matrix(c(
  3.0, 0.01, 0.01, 0.01, 0.01,
  0.01, 3.0, 0.01, 0.01, 0.01,
  0.01, 0.01, 3.0, 0.01, 0.01,
  0.01, 0.01, 0.01, 3.0, 0.01,
  0.01, 0.01, 0.01, 0.01, 3.0
), 5, 5, byrow = TRUE)
```

## Random seed

A random number is chosen 1-5 (10 times) to determine which population receives the infection seed

```{r seed, echo=FALSE}
seed_pop <- sample(1:5, 10, replace=TRUE)
```

## SIR model

```{r SIR model}
sir <- odin({
  
  # Equations for transitions between compartments by group
  
  deriv(S[]) <- (gamma*N[i]) - (lambda[i]*S[i]) - (mu*S[i]) - seed[i]
  
  deriv(E[]) <- (lambda[i]*S[i]) - (theta*E[i]) - (mu*E[i])
  
  deriv(I[]) <- (theta*E[i]) - (mu*I[i]) - (sigma*I[i]) + seed[i]
  
  deriv (R[]) <- (sigma*I[i]) - (mu*R[i])
  
  # with more populations it became easier to separate lambda
  
  lambda[] <- briere * (
    beta[i,1] * I[1]/N[1] +
      beta[i,2] * I[2]/N[2] +
      beta[i,3] * I[3]/N[3] +
      beta[i,4] * I[4]/N[4] +
      beta[i,5] * I[5]/N[5]
  )
  
  beta <- parameter()
  
  
  # initial conditions
  
  N[] <- S[i] + E[i] + I[i] + R[i]
  
  initial(S[]) <- 600 
  
  initial(E[]) <- 0
  
  initial(I[]) <- 0 
  
  initial(R[]) <- 0
  
  # infection seeded on day 60 each year
  year <- floor(time/365) + 1
  seed[] <- if(
    i == seed_pop[year] &&
    time%%365 >= 60 &&
    time%%365 < 61
  ) I0 else 0
  
  # User defined parameters:
  
  briere <- interpolate(briere_time, briere_value, "linear")
  
  briere_time <- parameter(constant = TRUE)
  briere_value <- parameter(constant = TRUE)
  
  seed_pop <- parameter(constant = TRUE)
  
  I0 <- parameter(10) # seed is 10 infected birds
  
  mu <- parameter(0.0015) # sensitivity analysis 0.00382
  
  theta <- parameter(0.5)
  
  sigma <- parameter(0.57) # house sparrow
  # sensitivity analysis: 0.2
  
  # birth season may-july
  gamma <- if (time %% 365 >= 121 && time %% 365 <= 213) 0.0059 else 0 #0.5 births per bird
  # sensitivity analysis  0.015 (1.5 births per bird)
  
  # Dimensions of arrays
  
  dim(S) <- 5
  
  dim(E) <- 5
  
  dim(I) <- 5
  
  dim(R) <- 5
  
  dim(N) <- 5
  
  dim(beta) <- c(5, 5)
  
  dim(lambda) <- 5
  
  dim(briere_time, briere_value) <- parameter(rank = 1)
  
  dim(seed_pop) <- 10
  
  dim(seed) <- 10
})

```

## Including average daily temperature

Temperatures downloaded from <https://meteostat.net/en/place/it/bologna?s=16140&t=2025-01-01/2025-12-31>

```{r briere function, echo=TRUE}
#daily_temperatures <- read.csv("~/bologna temperature series.csv")
#bologna.temperature.series <- read.csv("C:/Users/nsgoo/OneDrive/Desktop/Research Project/bologna temperature series.csv")

bologna.temperature.series <- read.csv("C:/Users/nsgoo/OneDrive/Desktop/Research Project/bologna temperature series.csv")

daily_temperatures <- bologna.temperature.series

daily_temp <- daily_temperatures$tavg
str(daily_temp)

BriereNorm <- function(temp, T_minB, T_maxB, mB) {
  
  briere <- (temp - T_minB) * (T_maxB - temp)^mB
  
  briere[briere < 0] <- 0
  briere[is.na(briere)] <- 0
  
  briere <- briere / max(briere, na.rm = TRUE)
  
  return(briere)
}


briere_value <- rep(BriereNorm(
  temp = daily_temp, 
  T_minB = 14, 
  T_maxB = 34.3, 
  mB = 0.7), #24 peak
  10
)

briere_time <- seq(0, length(briere_value)-1, by=1)

temp_seq <- seq(0.3, 31.15, by = 0.1)


plot(temp_seq,
     BriereNorm(temp_seq, 14, 34.3, 0.7), 
     type = "l",
     xlab = "Temperature (°C)",
     ylab = "Brière response")
```

## Simulation

```{r simulation}
sys <- dust_system_create(
  sir,
  list(
    beta = beta,
    briere_time = briere_time,
    briere_value = briere_value,
    seed_pop=seed_pop
  )
)

dust_system_set_state_initial(sys)

t <- seq(0, 3649) # 5 years

y <- dust_system_simulate(sys, t)

y <- dust_unpack_state(sys, y)

```

## Plot of infected birds

```{r infected birds, echo=TRUE}
df <- as.data.frame(t(y$I))
df$t <- t
colnames(df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")
df$Total <- rowSums(df[, c("P1", "P2", "P3", "P4", "P5")])
df_long <- pivot_longer(df, cols = c("P1", "P2", "P3", "P4", "P5", "Total"),
                        names_to = "Population",
                        values_to = "I")

ggplot(df_long, aes(x = t, y = I, colour = Population)) +
  geom_line(data = subset(df_long, Population != "Total")
  ) +
  labs(
    title = "Simulated infection dynamics across five avian populations (10 years)",
    y = "Infected birds",
    x = "Time (days)"
  ) +
  theme_minimal()


# infected plot by year

month_breaks <- c(0,31,59,90,120,151,181,212,243,273,304,334,365)

ggplot(data = subset(df_long, Population != "Total"), aes(x = t %% 365, y = I, color = Population)) +
  geom_line() +
  facet_wrap(~(floor(t/365)+1), ncol = 1) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = c("Jan","Feb","Mar","Apr","May","Jun",
               "Jul","Aug","Sep","Oct","Nov","Dec","Jan")
  ) +
  labs(
    x = "Month",
    y = "Infected birds",
    title = "Simulated infection dynamics across five avian populations (10 years)"
  ) +
  theme_minimal()

## Summary table
summary_table <- df_long |>
  dplyr::group_by(Population) |>
  dplyr::summarise(
    Peak = max(I),
    Peak_day = t[which.max(I)],
    .groups = "drop"
  )
```

## Plot of recovered birds

```{r recovered birds, echo=TRUE}
R_df <- as.data.frame(t(y$R))
R_df$t <- t
colnames(R_df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")

R_long <- pivot_longer(
  R_df,
  cols = c("P1", "P2", "P3", "P4", "P5"),
  names_to = "Population",
  values_to = "R"
)

ggplot(R_long, aes(x = t, y = R, color = Population)) +
  geom_line() +
  labs(title = "Simulated recovered birds in 5 populations over 10 years", 
       x = "Time (days)",
       y = "Recovered") +
  theme_minimal()
```

## Plot of susceptible birds

```{r susceptible birds, echo=TRUE}
S_df <- as.data.frame(t(y$S))
S_df$t <- t

colnames(S_df)[1:5] <- c(
  "P1",
  "P2",
  "P3",
  "P4",
  "P5"
)

S_long <- pivot_longer(
  S_df,
  cols = c("P1","P2","P3","P4","P5"),
  names_to = "Population",
  values_to = "S"
)

ggplot(S_long,
       aes(t, S, colour = Population)) +
  geom_line()+
  labs(
    title = "Simulated susceptible population size in 5 avian populations over 10 years",
    x = "Time (days)",
    y = "Number of Susceptible Birds") +
  theme_minimal()
```

## Plot of total birds

```{r total birds, echo=TRUE}
N_mat <- y$S + y$E + y$I + y$R

N_df <- as.data.frame(t(N_mat))
N_df$t <- t

colnames(N_df)[1:5] <- c(
  "population 1",
  "population 2",
  "population 3",
  "population 4",
  "population 5"
)

N_long <- pivot_longer(
  N_df,
  cols = c("population 1",
           "population 2",
           "population 3",
           "population 4",
           "population 5"),
  names_to = "group",
  values_to = "N"
)

ggplot(N_long, aes(x = t, y = N, color = group)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Total population size over time",
    x = "Time (days)",
    y = "Number of birds"
  ) +
  theme_minimal()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
