---
title: "09-07 model for scenarios"
output: html_document
date: "2026-07-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Equations

$$
\frac{\mathrm{d}S_i}{\mathrm{d}t} = (\gamma N_i) - (\lambda S_i) - (\mu S_i)
$$ $$
\frac{\mathrm{d}E_i}{\mathrm{d}t} = (\lambda S_i)  - (\theta E_i) - (\mu E_i)
$$ $$
\frac{\mathrm{d}I_i}{\mathrm{d}t} = (\theta E_i) - (\mu I_i) - (\sigma I_i) + seed
$$ $$
\frac{\mathrm{d}R_i}{\mathrm{d}t} = (\sigma I_i) - (\mu R_i)
$$ $$
\lambda = briere(beta_{i,1} \frac {I_1}{N_i} + beta_{i,2} \frac {I_2}{N_i} + beta_{i,3} \frac {I_3}{N_i}+ beta_{i,4} \frac {I_4}{N_i}+ beta_{i,5} \frac {I_5}{N_i})
$$

## Matrix
This represents the different transmission rates between 5 avian populations

$$
\begin{aligned}
beta &=
\begin{bmatrix}
\beta_{11} & \beta_{12} & \beta_{13} & \beta_{14} & \beta_{15}\\
\beta_{21} & \beta_{22} & \beta_{23} & \beta_{24} & \beta_{25}\\
\beta_{31} & \beta_{32} & \beta_{33} & \beta_{34} & \beta_{35}\\
\beta_{41} & \beta_{42} & \beta_{43} & \beta_{44} & \beta_{45}\\
\beta_{51} & \beta_{52} & \beta_{53} & \beta_{54} & \beta_{55}
\end{bmatrix}
\end{aligned}
$$
```{r matrix}
library(odin2)
library(dust2)
library(monty)
library(ggplot2)
library(tidyr)
library(patchwork)

# Scenario 1
beta <- matrix(c(
  0.5, 0, 0, 0, 0,
  0, 0.5, 0, 0, 0,
  0, 0, 0.5, 0, 0,
  0, 0, 0, 0.5, 0,
  0, 0, 0, 0, 0.5
), 5, 5, byrow = TRUE)

# Scenario 2
beta <- matrix(c(
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5
), 5, 5, byrow = TRUE)

# Scenario 3
beta <- matrix(c(
  0.5, 0.05, 0.05, 0.05, 0.05,
  0.05, 0.5, 0.05, 0.05, 0.05,
  0.05, 0.05, 0.5, 0.05, 0.05,
  0.05, 0.05, 0.05, 0.5, 0.05,
  0.05, 0.05, 0.05, 0.05, 0.5
), 5, 5, byrow = TRUE)

# Scenario 4
beta <- matrix(c(
  0.5, 0.1, 0.1, 0.1, 0.1,
  0.1, 0.5, 0.1, 0.1, 0.1,
  0.1, 0.1, 0.5, 0.1, 0.1,
  0.1, 0.1, 0.1, 0.5, 0.1,
  0.1, 0.1, 0.1, 0.1, 0.5
), 5, 5, byrow = TRUE)

# Scenario 5
beta <- matrix(c(
  0.3, 0.3, 0.3, 0.3, 0.3,
  0.3, 0.3, 0.3, 0.3, 0.3,
  0.3, 0.3, 0.3, 0.3, 0.3,
  0.3, 0.3, 0.3, 0.3, 0.3,
  0.3, 0.3, 0.3, 0.3, 0.3
), 5, 5, byrow = TRUE)

# Scenario 6
beta <- matrix(c(
  0.3, 0.03, 0.03, 0.03, 0.03,
  0.03, 0.3, 0.03, 0.03, 0.03,
  0.03, 0.03, 0.3, 0.03, 0.03,
  0.03, 0.03, 0.03, 0.3, 0.03,
  0.03, 0.03, 0.03, 0.03, 0.3
), 5, 5, byrow = TRUE)

# Scenario 7
beta <- matrix(c(
  0.3, 0.06, 0.06, 0.06, 0.06,
  0.06, 0.3, 0.06, 0.06, 0.06,
  0.06, 0.06, 0.3, 0.06, 0.06,
  0.06, 0.06, 0.06, 0.3, 0.06,
  0.06, 0.06, 0.06, 0.06, 0.3
), 5, 5, byrow = TRUE)

# Scenario 8
beta <- matrix(c(
  0.7, 0.7, 0.7, 0.7, 0.7,
  0.7, 0.7, 0.7, 0.7, 0.7,
  0.7, 0.7, 0.7, 0.7, 0.7,
  0.7, 0.7, 0.7, 0.7, 0.7,
  0.7, 0.7, 0.7, 0.7, 0.7
), 5, 5, byrow = TRUE)

# Scenario 9
beta <- matrix(c(
  0.7, 0.07, 0.07, 0.07, 0.07,
  0.07, 0.7, 0.07, 0.07, 0.07,
  0.07, 0.07, 0.7, 0.07, 0.07,
  0.07, 0.07, 0.07, 0.7, 0.07,
  0.07, 0.07, 0.07, 0.07, 0.7
), 5, 5, byrow = TRUE)

# Scenario 10
beta <- matrix(c(
  0.7, 0.14, 0.14, 0.14, 0.14,
  0.14, 0.7, 0.14, 0.14, 0.14,
  0.14, 0.14, 0.7, 0.14, 0.14,
  0.14, 0.14, 0.14, 0.7, 0.14,
  0.14, 0.14, 0.14, 0.14, 0.7
), 5, 5, byrow = TRUE)

# Scenario 11
beta <- matrix(c(
  0.50, 0.10, 0.08, 0.02, 0.01,
  0.10, 0.50, 0.10, 0.01, 0.01,
  0.08, 0.10, 0.50, 0.10, 0.08,
  0.02, 0.01, 0.10, 0.50, 0.10,
  0.01, 0.01, 0.08, 0.10, 0.50
), nrow = 5, byrow = TRUE)

# Scenario 12
beta <- matrix(c(
  0.50, 0.02, 0.10, 0.02, 0.02,
  0.02, 0.50, 0.10, 0.02, 0.02,
  0.10, 0.10, 0.50, 0.10, 0.10,
  0.02, 0.02, 0.10, 0.50, 0.02,
  0.02, 0.02, 0.10, 0.02, 0.50
), nrow = 5, byrow = TRUE)

# Scenario 13
beta <- matrix(c(
  0.50, 0.1, 0.01, 0.05, 0.06,
  0.1, 0.50, 0.06, 0.05, 0.05,
  0.01, 0.06, 0.50, 0.02, 0.01,
  0.05, 0.05, 0.02, 0.50, 0.08,
  0.06, 0.05, 0.01, 0.08, 0.50
), nrow = 5, byrow = TRUE)
```

## Random seed

A random number is chosen 1-5 (10 times) to determine which population receives the infection seed

```{r seed, echo=FALSE}
seed_pop <- sample(1:5, 10, replace=TRUE)
```

## Temperature

Bologna temperatures series downloaded from <https://meteostat.net/en/place/it/bologna?s=16140&t=2025-01-01/2025-12-31>

```{r temperature, echo=FALSE}
bologna.temperature.series <- read.csv("C:/Users/nsgoo/OneDrive/Desktop/Research Project/bologna temperature series.csv")
daily_temperatures <- bologna.temperature.series

daily_temp <- daily_temperatures$tavg
str(daily_temp)

BriereNorm <- function(temp, T_minB, T_maxB, mB) {
  
  briere <- (temp - T_minB) * (T_maxB - temp)^mB
  
  briere[briere < 0] <- 0
  briere[is.na(briere)] <- 0
  
  briere <- briere / max(briere, na.rm = TRUE)
  
  return(briere)
}


briere_value <- rep(BriereNorm(
  temp = daily_temp, 
  T_minB = 14, 
  T_maxB = 34.3, 
  mB = 0.7), #24 peak
  10
)

briere_time <- seq(0, length(briere_value)-1, by=1)

temp_seq <- seq(0.3, 31.15, by = 0.1)


plot(temp_seq,
     BriereNorm(temp_seq, 14, 34.3, 0.7), 
     type = "l",
     xlab = "Temperature (°C)",
     ylab = "Briere response")
```

## SIR model

```{r SIR model}
sir <- odin({
  
  # Equations for transitions between compartments by group
  
  deriv(S[]) <- (gamma[i]*N[i]) - (lambda[i]*S[i]) - (mu*S[i]) - seed[i]
  
  deriv(E[]) <- (lambda[i]*S[i]) - (theta*E[i]) - (mu*E[i])
  
  deriv(I[]) <- (theta*E[i]) - (mu*I[i]) - (sigma*I[i]) + seed[i]
  
  deriv (R[]) <- (sigma*I[i]) - (mu*R[i])
  
  lambda[] <- briere * (
    beta[i,1] * I[1]/N[i] +
      beta[i,2] * I[2]/N[i] +
      beta[i,3] * I[3]/N[i] +
      beta[i,4] * I[4]/N[i] +
      beta[i,5] * I[5]/N[i]
  )
  
  beta <- parameter()
  
  
  # initial conditions
  
  N[] <- S[i] + E[i] + I[i] + R[i]
  
  initial(S[]) <- 500 
  
  initial(E[]) <- 0
  
  initial(I[]) <- 0 
  
  initial(R[]) <- 0
  
  # seed in march every year
  year <- floor(time/365) + 1
  seed[] <- if(
    i == seed_pop[year] &&
    time%%365 >= 60 && #182 for July
    time%%365 < 61 #183 for July
  ) I0 else 0
  
  # User defined parameters:
  
  briere <- interpolate(briere_time, briere_value, "linear")
  
  briere_time <- parameter(constant = TRUE)
  briere_value <- parameter(constant = TRUE)
  
  seed_pop <- parameter(constant = TRUE)
  
  I0 <- parameter(1) # initial infected
  
  sigma <- parameter(0.25) # recovery rate
  
  mu <- parameter(0.000913242) # death rate (1/(365x3)), 3 year life span 
  # mu <- parameter(0.000548) = 5 YEARS 
  
  theta <- parameter(0.5) # incubation period

  # birth rate
  gamma[] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0027548209 else 0 # 1/ (121x3)
  # replenishes deaths
  # gamma[] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0016528926 else 0 = 5 YEARS 
  
  
  # Dimensions of arrays
  
  dim(S) <- 5
  
  dim(E) <- 5
  
  dim(I) <- 5
  
  dim(R) <- 5
  
  dim(N) <- 5
  
  
  dim(beta) <- c(5, 5)
  
  dim(briere_time, briere_value) <- parameter(rank = 1)
  
  dim(lambda) <- 5
  
  dim(seed_pop) <- 10
  
  dim(seed) <- 10
  
  dim(gamma) <- 5
})
```
## Simulation

```{r simulation}
sys <- dust_system_create(
  sir,
  list(
    beta = beta,
    briere_time = briere_time,
    briere_value = briere_value,
    seed_pop=seed_pop
  )
)

dust_system_set_state_initial(sys)

t <- seq(0, 3649) # 10 years 3649

y <- dust_system_simulate(sys, t)

y <- dust_unpack_state(sys, y)
```

## SIR model with different recovery rates in each population
```{r recovery rates, echo=TRUE}
sigma = c(
  0.5,  # population 1
  0.25,  # population 2
  0.17,  # population 3
  0.2,  # population 4
  0.33   # population 5
)

# Scenario 14
beta <- matrix(c(
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5
), 5, 5, byrow = TRUE)

# Scenario 15
beta <- matrix(c(
  0.5, 0.05, 0.05, 0.05, 0.05,
  0.05, 0.5, 0.05, 0.05, 0.05,
  0.05, 0.05, 0.5, 0.05, 0.05,
  0.05, 0.05, 0.05, 0.5, 0.05,
  0.05, 0.05, 0.05, 0.05, 0.5
), 5, 5, byrow = TRUE)

# Scenario 16
beta <- matrix(c(
  0.5, 0.1, 0.1, 0.1, 0.1,
  0.1, 0.5, 0.1, 0.1, 0.1,
  0.1, 0.1, 0.5, 0.1, 0.1,
  0.1, 0.1, 0.1, 0.5, 0.1,
  0.1, 0.1, 0.1, 0.1, 0.5
), 5, 5, byrow = TRUE)

# Scenario 17
beta <- matrix(c(
  0.50, 0.1, 0.01, 0.05, 0.06,
  0.1, 0.50, 0.06, 0.05, 0.05,
  0.01, 0.06, 0.50, 0.02, 0.01,
  0.05, 0.05, 0.02, 0.50, 0.08,
  0.06, 0.05, 0.01, 0.08, 0.50
), nrow = 5, byrow = TRUE)

sir <- odin({
  
  # Equations for transitions between compartments by group
  
  deriv(S[]) <- (gamma[i]*N[i]) - (lambda[i]*S[i]) - (mu*S[i]) - seed[i]
  
  deriv(E[]) <- (lambda[i]*S[i]) - (theta*E[i]) - (mu*E[i])
  
  deriv(I[]) <- (theta*E[i]) - (mu*I[i]) - (sigma[i]*I[i]) + seed[i]
  
  deriv (R[]) <- (sigma[i]*I[i]) - (mu*R[i])
  
  lambda[] <- briere * (
    beta[i,1] * I[1]/N[1] +
      beta[i,2] * I[2]/N[2] +
      beta[i,3] * I[3]/N[3] +
      beta[i,4] * I[4]/N[4] +
      beta[i,5] * I[5]/N[5]
  )
  
  beta <- parameter()
  
  
  # initial conditions
  
  N[] <- S[i] + E[i] + I[i] + R[i]
  
  initial(S[]) <- 300 
  
  initial(E[]) <- 0
  
  initial(I[]) <- 0 
  
  initial(R[]) <- 0
  
  # seed in march every year
  year <- floor(time/365) + 1
  seed[] <- if(
    i == seed_pop[year] &&
    time%%365 >= 60 && 
    time%%365 < 61 
  ) I0 else 0
  
  # User defined parameters:
  
  briere <- interpolate(briere_time, briere_value, "linear")
  
  briere_time <- parameter(constant = TRUE)
  briere_value <- parameter(constant = TRUE)
  
  seed_pop <- parameter(constant = TRUE)
  
  I0 <- parameter(1) # initial infected
  
  mu <- parameter(0.000913242) # death rate (1/(365x3)), 3 year life span 
  # 5 years = 0.000548
  
  theta <- parameter(0.5) # incubation period
  
  # birth rate
  gamma[] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0027548209 else 0 # 1/ (121x3)
  # replenishes deaths
  # 5 years = 0.0016528926
  
  sigma <- parameter()
  
  # Dimensions of arrays
  
  dim(S) <- 5
  
  dim(E) <- 5
  
  dim(I) <- 5
  
  dim(R) <- 5
  
  dim(N) <- 5
  
  
  dim(beta) <- c(5, 5)
  
  dim(briere_time, briere_value) <- parameter(rank = 1)
  
  dim(lambda) <- 5
  
  dim(seed_pop) <- 10
  
  dim(seed) <- 10
  
  dim(sigma) <- 5
  
  dim(gamma) <- 5
})

##############
# Simulation #
##############

sys <- dust_system_create(
  sir,
  list(
    beta = beta,
    briere_time = briere_time,
    briere_value = briere_value,
    seed_pop=seed_pop,
    sigma=sigma
  )
)

dust_system_set_state_initial(sys)

t <- seq(0, 3649) # 10 years 3649

y <- dust_system_simulate(sys, t)

y <- dust_unpack_state(sys, y)
```
## Plots of infected birds

```{r infected birds, echo=TRUE}

# data frames

df <- as.data.frame(t(y$I))
df$t <- t
colnames(df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")
df_long <- pivot_longer(df, cols = c("P1", "P2", "P3", "P4", "P5"),
                        names_to = "Population",
                        values_to = "I")


prop <- y$I / (y$S + y$E + y$I + y$R) # proportion infected

df <- as.data.frame(t(prop))
df$t <- t
colnames(df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")

df_propI_long <- pivot_longer(df,
                              cols = c(P1, P2, P3, P4, P5),
                              names_to = "Population",
                              values_to = "PropI")

df_plotI <- df_propI_long[df_propI_long$t >= (365 * 4), ]  # last 6 years

## 6 years ##

month_breaks <- c(0,31,59,90,120,151,181,212,243,273,304,334)

plotI6 <- ggplot(data = subset(df_plotI), aes(x = t %% 365, y = PropI, color = Population)) +
  geom_line() +
  facet_wrap(~(floor(t/365)+1), ncol = 1) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = c("Jan","Feb","Mar","Apr","May","Jun",
               "Jul","Aug","Sep","Oct","Nov","Dec"),
    expand = c(0, 0)
  ) +
  labs(
    x = NULL,
    y = "Proportion Infected"
  ) +
  theme_minimal()+
  theme(
    axis.text.y = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )


## 10 years ##

plotI10 <- ggplot(data = subset(df_propI_long), aes(x = t %% 365, y = PropI, color = Population)) +
  geom_line() +
  facet_wrap(~(floor(t/365)+1), ncol = 1) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = c("Jan","Feb","Mar","Apr","May","Jun",
               "Jul","Aug","Sep","Oct","Nov","Dec")
  ) +
  labs(
    x = NULL,
    y = "Proportion Infected"
  ) +
  theme_minimal()+
  theme(
    axis.text.y = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )
```
## Plots of recovered birds

```{r recovered birds, echo=TRUE}
# dataframes 

R_df <- as.data.frame(t(y$R))
R_df$t <- t
colnames(R_df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")

R_long <- pivot_longer(
  R_df,
  cols = c("P1", "P2", "P3", "P4", "P5"),
  names_to = "Population",
  values_to = "R"
)

propR <- y$R / (y$S + y$E + y$I + y$R) # proportion recovered

df <- as.data.frame(t(propR))
df$t <- t
colnames(df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")

df_propR_long <- pivot_longer(df,
                              cols = c(P1, P2, P3, P4, P5),
                              names_to = "Population",
                              values_to = "PropR")

df_plotR <- df_propR_long[df_propR_long$t >= (365 * 4), ]  # last 6 years

## 6 years ##

month_breaks <- (0:71) * (365/12)

plotR6 <- ggplot(df_plotR, aes(x = t - 365*4, y = PropR, color = Population)) +
  geom_rect(
    data = data.frame(
      xmin = 365 * 0:5,
      xmax = 365 * 1:6,
      fill = rep(c("grey90", "white"), length.out = 6)
    ),
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  scale_fill_identity() +
  geom_line() +
  labs(x= "Time",
       y = " Proportion Recovered") +
  scale_x_continuous(
    breaks = seq(0, 365*6, by = 365/12),   # monthly lines
    labels = function(x) ifelse(x %% 365 == 0, (x / 365) + 4, ""),  # only years
    name = "Year"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(colour = "grey80", linewidth = 0.15),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )


## 10 years ##

month_breaks <- (0:119) * (365/12)

plotR10 <- ggplot(df_propR_long, aes(x = t, y = PropR, color = Population)) +
  geom_rect(
    data = data.frame(
      xmin = 365 * 0:9,
      xmax = 365 * 1:10,
      fill = rep(c("grey90", "white"), length.out = 10)
    ),
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  scale_fill_identity() +
  geom_line() +
  labs(x = "Time",
       y = " Proportion Recovered") +
  scale_x_continuous(
    breaks = seq(0, 365*10, by = 365/12),   # monthly lines
    labels = function(x) ifelse(x %% 365 == 0, (x / 365) + 4, ""),  # only years
    name = "Year"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(colour = "grey80", linewidth = 0.15),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
  )
```

## Plots of susceptible birds

```{r susceptible birds, echo=TRUE}
# dataframes

S_df <- as.data.frame(t(y$S))
S_df$t <- t

colnames(S_df)[1:5] <- c(
  "P1",
  "P2",
  "P3",
  "P4",
  "P5"
)

S_long <- pivot_longer(
  S_df,
  cols = c("P1","P2","P3","P4","P5"),
  names_to = "Population",
  values_to = "S"
)

propS <- y$S / (y$S + y$E + y$I + y$R) # proportion susceptible

df <- as.data.frame(t(propS))
df$t <- t
colnames(df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")

df_propS_long <- pivot_longer(df,
                              cols = c(P1, P2, P3, P4, P5),
                              names_to = "Population",
                              values_to = "PropS")

df_plotS <- df_propS_long[df_propS_long$t >= (365 * 4), ]  # last 6 years

## 6 years ## 

month_breaks <- (0:71) * (365/12)

plotS6 <- ggplot(df_plotS, aes(x = t - 365*4, y = PropS, color = Population)) +
  geom_rect(
    data = data.frame(
      xmin = 365 * 0:5,
      xmax = 365 * 1:6,
      fill = rep(c("grey90", "white"), length.out = 6)
    ),
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  scale_fill_identity() +
  geom_line() +
  labs(x = "Time",
       y = " Proportion Susceptible") +
  scale_x_continuous(
    breaks = seq(0, 365*6, by = 365/12),   # monthly lines
    labels = function(x) ifelse(x %% 365 == 0, (x / 365) + 4, ""),  # only years
    name = "Year"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(colour = "grey80", linewidth = 0.15),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

## 10 years ##

month_breaks <- (0:119) * (365/12)

plotS10 <- ggplot(df_propS_long, aes(x = t, y = PropS, color = Population)) +
  geom_rect(
    data = data.frame(
      xmin = 365 * 0:9,
      xmax = 365 * 1:10,
      fill = rep(c("grey90", "white"), length.out = 10)
    ),
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  scale_fill_identity() +
  geom_line() +
  labs(x = "Time",
       y = " Proportion Susceptible") +
  scale_x_continuous(
    breaks = seq(0, 365*10, by = 365/12),   # monthly lines
    labels = function(x) ifelse(x %% 365 == 0, (x / 365) + 4, ""),  # only years
    name = "Year"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(colour = "grey80", linewidth = 0.15),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

## Total birds
```{r total birds, echo=TRUE}
N_mat <- y$S + y$E + y$I + y$R

N_df <- as.data.frame(t(N_mat))
N_df$t <- t

colnames(N_df)[1:5] <- c(
  "population 1",
  "population 2",
  "population 3",
  "population 4",
  "population 5"
)

N_long <- pivot_longer(
  N_df,
  cols = c("population 1",
           "population 2",
           "population 3",
           "population 4",
           "population 5"),
  names_to = "group",
  values_to = "N"
)

ggplot(N_long, aes(x = t, y = N, color = group)) +
  geom_rect(
    data = data.frame(
      xmin = 365 * 0:9,
      xmax = 365 * 1:10,
      fill = rep(c("grey90", "white"), length.out = 10)
    ),
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill),
    inherit.aes = FALSE,
    alpha = 0.3
  ) +
  scale_fill_identity() +
  geom_line(linewidth = 1) +
  labs(
    title = "Total population size over time",
    x = "Time (days)",
    y = "Number of birds"
  ) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = rep(1:12, 10),
    name = "Month"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(colour = "grey80", linewidth = 0.15),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

## Panel
```{r panel, echo=TRUE}
(plotI6) /
  (plotR6 | plotS6) +
  plot_annotation(tag_levels = "a") +
  plot_layout(heights = c(2, 1))
```
