---
title: "16-06 update"
output: html_document
date: "2026-06-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Modifications

This version has balanced population dynamics, standardised temperature using briere function and includes the bird breeding season.
1 infected bird is imported into each population in March of each year.
Simulation runs over 5 years.

## Equations

$$
\frac{\mathrm{d}S_i}{\mathrm{d}t} = (\gamma N_i) - (briere(m_{i,1} \frac {I_1}{N_1} + m_{i,2} \frac {I_2}{N_2} + m_{i,3} \frac {I_3}{N_3})S_i) - (\mu S_i)
$$ $$
\frac{\mathrm{d}E_i}{\mathrm{d}t} = (briere(m_{i,1} \frac {I_1}{N_1} + m_{i,2} \frac {I_2}{N_2} + m_{i,3} \frac {I_3}{N_3})S_i) - (\theta E_i) - (\mu E_i)
$$ $$
\frac{\mathrm{d}I_i}{\mathrm{d}t} = (\theta E_i) - (\mu I_i) - (\sigma I_i) + seed
$$ $$
\frac{\mathrm{d}R_i}{\mathrm{d}t} = (\sigma I_i) - (\mu R_i)
$$

## Matrix

$$
\begin{aligned}
m &=
\begin{bmatrix}
0.9 & 0.15 & 0.02 \\
0.15 & 0.55 & 0.08 \\
0.02 & 0.08 & 0.25
\end{bmatrix}
\qquad
m &=
\begin{bmatrix}
\beta_{11} & \beta_{12} & \beta_{13}\\
\beta_{21} & \beta_{22} & \beta_{23}\\
\beta_{31} & \beta_{32} & \beta_{33}
\end{bmatrix}
\end{aligned}
$$
Population 1 has high competence, P2 intermediate and P3 low
P1 and P2 are nearby, P2 and P3 are moderately separated and P3 and P1 are far apart

```{r matrix}
library(odin2)
library(dust2)
library(monty)
library(ggplot2)
library(tidyr)

m <- matrix(c(0.9,0.15,0.02,0.15,0.55,0.08,0.02,0.08,0.25), 3, 3, byrow=TRUE)  # beta matrix of 3 populations
```
## SIR model
```{r SIR model}
sir <- odin({
  
  # Equations for transitions between compartments by group
  
  deriv(S[]) <- (gamma*N[i]) - (briere*(m[i,1] * I[1]/N[1] + m[i,2] * I[2]/N[2] + m[i,3] * I[3]/N[3])*S[i]) - (mu*S[i]) - seed
  
  deriv(E[]) <- (briere*(m[i,1] * I[1]/N[1] + m[i,2] * I[2]/N[2] + m[i,3] * I[3]/N[3])*S[i]) - (theta*E[i]) - (mu*E[i])
  
  deriv(I[]) <- (theta*E[i]) - (mu*I[i]) - (sigma*I[i]) + seed #seed imports 1 infection every march
  
  deriv (R[]) <- (sigma*I[i]) - (mu*R[i])
  
  m <- parameter()
  
  
  # initial conditions
  
  N[] <- S[i] + E[i] + I[i] + R[i]
  
  initial(S[]) <- 500 - I0
  
  initial(E[]) <- 0
  
  initial(I[]) <- 0 
  
  initial(R[]) <- 0
  
  
  seed <- if(time%%365>=60 && time%%365<61)I0 else 0 #if march 1st I0 = 1
  
  # User defined parameters:
  
  briere <- interpolate(briere_time, briere_value, "linear") # temperature function
  briere_time <- parameter(constant = TRUE)
  briere_value <- parameter(constant = TRUE)
  
  I0 <- parameter(1)
  
  mu <- parameter(0.0015) 
  
  theta <- parameter(0.5)
  
  sigma <- parameter(0.57)
  
  #gamma <- parameter(0.0015) # tested with constant gamma
  gamma <- if (time %% 365 >= 121 && time %% 365 <= 213) 0.0059 else 0 #(mu*365)/93 #may june july (93 days)
  
  # Dimensions of arrays
  
  dim(S) <- 3
  
  dim(E) <- 3
  
  dim(I) <- 3
  
  dim(R) <- 3
  
  dim(N) <- 3
  
  dim(m) <- c(3, 3)
  
  dim(briere_time, briere_value) <- parameter(rank = 1)
})

```

## Including average daily temperature

Temperatures downloaded from teams chat

```{r briere function, echo=TRUE}
daily_temperatures_MO <- read.csv("~/daily_temperatures.csv") # imported dataset (.csv) and only included march-october 
daily_temp <- daily_temperatures_MO$DAILY_AVG 

BriereNorm <- function(temp, T_minB, T_maxB, mB) {
  
  briere <- (temp - T_minB) * (T_maxB - temp)^mB
  
  briere[briere < 0] <- 0
  briere[is.na(briere)] <- 0
  
  briere <- briere / max(briere, na.rm = TRUE)
  
  return(briere)
}


briere_value <- rep(BriereNorm(
  temp = daily_temp, 
  T_minB = 0.3,
  T_maxB = 31.15,
  mB = 0.3), # peak at 24 degrees
  5
)

briere_time <- seq(0, length(briere_value)-1, by=1)

temp_seq <- seq(0.3, 31.15, by = 0.1)

plot(temp_seq,
     BriereNorm(temp_seq, 0.3, 31.15, 0.5),
     type = "l",
     xlab = "Temperature (°C)",
     ylab = "Brière response")
```
## Simulation

```{r simulation}
sys <- dust_system_create(
  sir,
  list(
    m = m,
    briere_time = briere_time,
    briere_value = briere_value
  )
)

dust_system_set_state_initial(sys)

t <- seq(0, 1824) # 5 years

y <- dust_system_simulate(sys, t)

y <- dust_unpack_state(sys, y)


```
## Plot of infected birds

I also used a dataset with constant temperature to check that this gave the expected output (same as previous model without temperature)

```{r infected birds, echo=TRUE}
# plot

df <- as.data.frame(t(y$I))
df$t <- t
colnames(df)[1:3] <- c("population 1", "population 2", "population 3")
df_long <- pivot_longer(df, cols = c("population 1", "population 2", "population 3"),
                        names_to = "group",
                        values_to = "I")
ggplot(df_long, aes(x = t, y = I, color = group)) +
  geom_line() +
  labs(title = "Infected birds per population (5 year simulation)", y = "Infected birds") +
  theme_minimal()

# by year
month_breaks <- c(0,31,59,90,120,151,181,212,243,273,304,334,365)

ggplot(df_long, aes(x = t %% 365, y = I, color = group)) +
  geom_line() +
  facet_wrap(~(floor(t/365)+1), ncol = 1) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = c("Jan","Feb","Mar","Apr","May","Jun",
               "Jul","Aug","Sep","Oct","Nov","Dec","Jan")
  ) +
  labs(
    x = "Month",
    y = "Infected birds"
  ) +
  theme_minimal()
```
## Plot of recovered birds

```{r recovered birds, echo=TRUE}
R_df <- as.data.frame(t(y$R))
R_df$t <- t
colnames(R_df)[1:3] <- c("population 1", "population 2", "population 3")

R_long <- pivot_longer(
  R_df,
  cols = c("population 1", "population 2", "population 3"),
  names_to = "group",
  values_to = "R"
)

ggplot(R_long, aes(x = t, y = R, color = group)) +
  geom_line() +
  labs(title = "Recovered birds over time",
       x = "Time (days)",
       y = "Recovered (R)") +
  theme_minimal()
```
## Plot of total birds

```{r total birds, echo=TRUE}

N_mat <- y$S + y$E + y$I + y$R

N_df <- as.data.frame(t(N_mat))
N_df$t <- t

colnames(N_df)[1:3] <- c(
  "population 1",
  "population 2",
  "population 3"
)

N_long <- pivot_longer(
  N_df,
  cols = c("population 1",
           "population 2",
           "population 3"),
  names_to = "group",
  values_to = "N"
)

ggplot(N_long, aes(x = t, y = N, color = group)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Total population size over time",
    x = "Time (days)",
    y = "Number of birds"
  ) +
  theme_minimal()

```

## Plot of susceptible birds

```{r susceptible birds, echo=TRUE}

S_df <- as.data.frame(t(y$S))
S_df$t <- t

colnames(S_df)[1:3] <- c(
  "population 1",
  "population 2",
  "population 3"
)

S_long <- pivot_longer(
  S_df,
  cols = c("population 1","population 2","population 3"),
  names_to = "group",
  values_to = "S"
)

ggplot(S_long,
       aes(t, S, colour = group)) +
  geom_line()+
  theme_minimal()

```
