---
title: "Untitled"
output: html_document
date: "2026-08-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages



```{r packages}
library(odin2)
library(dust2)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Transmission matrices

You can also embed plots, for example:

```{r matrices, echo=FALSE}
# Scenario 1
beta <- matrix(c(
  0.5, 0, 0, 0, 0,
  0, 0.5, 0, 0, 0,
  0, 0, 0.5, 0, 0,
  0, 0, 0, 0.5, 0,
  0, 0, 0, 0, 0.5
), 5, 5, byrow = TRUE)

# Scenario 2 -  attack rate may be too high
beta <- matrix(c(
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5,
  0.5, 0.5, 0.3, 0.5, 0.5,
  0.5, 0.5, 0.5, 0.5, 0.5
), 5, 5, byrow = TRUE)

# Scenario 3
beta <- matrix(c(
  0.5, 0.05, 0.05, 0.05, 0.05,
  0.05, 0.5, 0.05, 0.05, 0.05,
  0.05, 0.05, 0.5, 0.05, 0.05,
  0.05, 0.05, 0.05, 0.5, 0.05,
  0.05, 0.05, 0.05, 0.05, 0.5
), 5, 5, byrow = TRUE)

# Scenario 4
beta <- matrix(c(
  0.5, 0.005, 0.005, 0.005, 0.005,
  0.005, 0.5, 0.005, 0.005, 0.005,
  0.005, 0.005, 0.5, 0.005, 0.005,
  0.005, 0.005, 0.005, 0.5, 0.005,
  0.005, 0.005, 0.005, 0.005, 0.5
), 5, 5, byrow = TRUE)

# Scenario 5 
beta <- matrix(c(
  0.5, 0.0005, 0.0005, 0.0005, 0.0005,
  0.0005, 0.5, 0.0005, 0.0005, 0.0005,
  0.0005, 0.0005, 0.5, 0.0005, 0.0005,
  0.0005, 0.0005, 0.0005, 0.5, 0.0005,
  0.0005, 0.0005, 0.0005, 0.0005, 0.5
), 5, 5, byrow = TRUE)


# Scenario 6 
beta <- matrix(c(
  0.50, 0.005, 0, 0, 0,
  0.005, 0.50, 0.001, 0, 0,
  0, 0.001, 0.50, 0.005, 0.005,
  0, 0, 0.005, 0.50, 0.005,
  0, 0, 0.005, 0.005, 0.50
), 5, 5, byrow = TRUE)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Temperature function
```{r temperature, echo=FALSE}
# same temperature in all populations
daily_temperatures <- bologna.temperature.series

# different temperature in each population
daily_temperatures1 <- bologna.temperature.series.1
daily_temperatures2 <- bologna.temperature.series.2
daily_temperatures3 <- bologna.temperature.series.3
daily_temperatures4 <- bologna.temperature.series.4
daily_temperatures5 <- bologna.temperature.series.5


# Bologna daily temperature plot

bologna.temperature.series$date <- as.Date(
  as.POSIXct(bologna.temperature.series$date, format = "%d/%m/%Y %H:%M")
)

daily_temp <- ggplot(bologna.temperature.series, aes(x = date, y = tavg)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  labs(
    title = NULL,
    x = NULL,
    y = "Temperature (°C)"
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b",
    expand = c(0, 0)
  ) +
  theme_minimal()+
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 9),
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    )
  )


BriereNorm <- function(temp, T_minB, T_maxB, mB) {
  briere <- (temp - T_minB) * (T_maxB - temp)^mB
  briere[briere < 0] <- 0
  briere[is.na(briere)] <- 0
  briere <- briere / max(briere, na.rm = TRUE)
  return(briere)
}

temp_seq <- seq(0.3, 31.15, by = 0.1)

# same temperature
daily_temp <- daily_temperatures$tavg

briere_value <- rep(BriereNorm(
  temp = daily_temp, 
  T_minB = 14, 
  T_maxB = 34.3, 
  mB = 0.7), #24 peak
  10
)

briere_time <- seq(0, length(briere_value)-1, by=1)

# different temperatures

# patch 1 briere
daily_temp1 <- daily_temperatures1$tavg

briere_value1 <- rep(BriereNorm(
  temp = daily_temp, 
  T_minB = 14, 
  T_maxB = 34.3, 
  mB = 0.7), #24 peak
  10
)

briere_time1 <- seq(0, length(briere_value1)-1, by=1)

# patch 2 briere
daily_temp2 <- daily_temperatures2$tavg

briere_value2 <- rep(BriereNorm(
  temp = daily_temp2, 
  T_minB = 14, 
  T_maxB = 34.3, 
  mB = 0.7), #24 peak
  10
)

briere_time2 <- seq(0, length(briere_value2)-1, by=1)

# patch 3 briere
daily_temp3 <- daily_temperatures3$tavg

briere_value3 <- rep(BriereNorm(
  temp = daily_temp3, 
  T_minB = 14, 
  T_maxB = 34.3, 
  mB = 0.7), #24 peak
  10
)

briere_time3 <- seq(0, length(briere_value3)-1, by=1)

# patch 4 briere
daily_temp4 <- daily_temperatures4$tavg

briere_value4 <- rep(BriereNorm(
  temp = daily_temp4, 
  T_minB = 14, 
  T_maxB = 34.3, 
  mB = 0.7), #24 peak
  10
)

briere_time4 <- seq(0, length(briere_value4)-1, by=1)

# patch 5 briere
daily_temp5 <- daily_temperatures5$tavg

briere_value5 <- rep(BriereNorm(
  temp = daily_temp5, 
  T_minB = 14, 
  T_maxB = 34.3, 
  mB = 0.7), #24 peak
  10
)

briere_time5 <- seq(0, length(briere_value5)-1, by=1)


# briere function plot

briere_df <- data.frame(
  temp = temp_seq,
  response = BriereNorm(temp_seq, 14, 34.3, 0.7)
)
briere_function <- ggplot(briere_df, aes(x = temp, y = response)) +
  geom_line(linewidth = 0.5) +
  labs(
    x = "Temperature (°C)",
    y = "Brière response"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 9)
  )


# temperature panel
(daily_temp | briere_function)+
  plot_annotation(tag_levels = "a")
```
## SIR model
delte/add # in appropriate sections for hetergeneous population scenarios
```{r sir model, echo=FALSE}

# run for scenarios 8-10
scenario_name <- "Size"
scenario_name <- "Dem"
scenario_name <- "Temp"

peak_results <- vector("list", 50) # for calculating mean peak and mean peak day for each scenario
trajectory_results <- vector("list", 50) # for figure 17

# Alternatively, use to compare infection prevalence of same seed pattern
seed_pop <- c(1, 5, 4, 3, 5, 3, 3, 1, 2, 4) # combination 1

for (rep in 1:50) {
  
  
  
  seed_pop <- sample(1:5, 10, replace=TRUE)
  
  
  #######
  # SIR #
  #######
  
  sir <- odin({
    
    # Equations for transitions between compartments by group
    
    # mu[i] for hetereogeneous demography scenario
    
    deriv(S[]) <- (gamma[i]*N[i]) - (lambda[i]*S[i]) - (mu*S[i]) - seed[i]
    
    deriv(E[]) <- (lambda[i]*S[i]) - (theta*E[i]) - (mu*E[i])
    
    deriv(I[]) <- (theta*E[i]) - (mu*I[i]) - (sigma*I[i]) + seed[i]
  
    deriv (R[]) <- (sigma*I[i]) - (mu*R[i])
    
    deriv(incidence[]) <- theta * E[i]
    
    # CONSTANT TEMPERATURE ACROSS PATCHES
    lambda[] <- 
      (beta[i,1] * briere) * I[1]/N[i] +
      (beta[i,2] * briere) * I[2]/N[i] +
      (beta[i,3] * briere) * I[3]/N[i] +
      (beta[i,4] * briere) * I[4]/N[i] +
      (beta[i,5] * briere) * I[5]/N[i]
    
    # DIFFERENT TEMPERATURES ACROSS PATCHES.
    
    #lambda[] <- 
    #  (beta[i,1] * briere1) * I[1]/N[i] +
    #  (beta[i,2] * briere2) * I[2]/N[i] +
    #  (beta[i,3] * briere3) * I[3]/N[i] +
    #  (beta[i,4] * briere4) * I[4]/N[i] +
    #  (beta[i,5] * briere5) * I[5]/N[i]
    
    beta <- parameter()
    
    
    # initial conditions
    
    N[] <- S[i] + E[i] + I[i] + R[i]
    
    # HOMOGENEOUS POPULATION SIZE
    
    initial(S[]) <- 500
    
    # DIFFERENT POPULATION SIZES
    
    #initial(S[1]) <- 500
    #initial(S[2]) <- 250 
    #initial(S[3]) <- 250 
    #initial(S[4]) <- 125 
    #initial(S[5]) <- 125 
    
    initial(E[]) <- 0
    
    initial(I[]) <- 0 
    
    initial(R[]) <- 0
    
    initial(incidence[]) <- 0
    
    # seed in july every year
    year <- floor(time/365) + 1
    seed[] <- if(
      i == seed_pop[year] &&
      time%%365 >= 182 && # July
      time%%365 < 183 
    ) 1 else 0
    
    # User defined parameters:
    
    # constant temperature
    
    briere <- interpolate(briere_time, briere_value, "linear")
    
    # variable temperature
    
    #briere1 <- interpolate(briere_time1, briere_value1, "linear")
    #briere2 <- interpolate(briere_time2, briere_value2, "linear")
    #briere3 <- interpolate(briere_time3, briere_value3, "linear")
    #briere4 <- interpolate(briere_time4, briere_value4, "linear")
    #briere5 <- interpolate(briere_time5, briere_value5, "linear")
    
    # constant temperature
    
    briere_time <- parameter(constant = TRUE)
    briere_value <- parameter(constant = TRUE)
    
    # variable temperature
    
    #briere_time1 <- parameter(constant = TRUE)
    #briere_time2 <- parameter(constant = TRUE)
    #briere_time3 <- parameter(constant = TRUE)
    #briere_time4 <- parameter(constant = TRUE)
    #briere_time5 <- parameter(constant = TRUE)
    #briere_value1 <- parameter(constant = TRUE)
    #briere_value2 <- parameter(constant = TRUE)
    #briere_value3 <- parameter(constant = TRUE)
    #briere_value4 <- parameter(constant = TRUE)
    #briere_value5 <- parameter(constant = TRUE)
    
    seed_pop <- parameter(constant = TRUE)
    
    
    sigma <- parameter(0.25) # recovery rate
    
    mu <- parameter(0.000913242) # death rate (1/(365x3)), 3 year life span 
    
    theta <- parameter(0.5) # incubation period
    
    # birth rate
    
    gamma[] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0027322404 else 0 # 1/ (122x3) 
    # replenishes deaths
    
    # HETEROGENEOUS DEMOGRAPHY SCENARIO
    
    #mu[1] <- 0.001369863 # 2 YEARS
    #mu[2] <- 0.000913242 # 3 YEARS 
    #mu[3] <- 0.0006849315 # 4 YEARS
    #mu[4] <- 0.0005479452 # 5 YEARS
    #mu[5] <- 0.000456621 # 6 YEARS
    
    #gamma[1] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0040983606 else 0 # 2 YEARS
    #gamma[2] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0027322404 else 0 # 3 YEARS
    #gamma[3] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0020491803 else 0 # 4 YEARS
    #gamma[4] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.00163934 else 0 # 5 YEARS
    #gamma[5] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0013661202 else 0 # 6 YEARS
    
    # Dimensions of arrays
    
    dim(S) <- 5
    
    dim(E) <- 5
    
    dim(I) <- 5
    
    dim(R) <- 5
    
    dim(N) <- 5
    
    dim(incidence) <- 5
    
    
    dim(beta) <- c(5, 5)
    
    # constant temperature
    
    dim(briere_time, briere_value) <- parameter(rank = 1)
    
    # variable temperature
    
    #dim(briere_time1, briere_value1) <- parameter(rank = 1)
    #dim(briere_time2, briere_value2) <- parameter(rank = 1)
    #dim(briere_time3, briere_value3) <- parameter(rank = 1)
    #dim(briere_time4, briere_value4) <- parameter(rank = 1)
    #dim(briere_time5, briere_value5) <- parameter(rank = 1)
    
    dim(lambda) <- 5
    
    dim(seed_pop) <- 10
    
    dim(seed) <- 10
    
    dim(gamma) <- 5
    
    #HETEROGENEOUS DEMOGRAPHY
    
    #dim(mu) <- 5
  })
  
  
  
  ##############
  # Simulation #
  ##############
  
  sys <- dust_system_create(
    sir,
    list(
      beta = beta,
      briere_time = briere_time, # constant temp
      briere_value = briere_value, # constant temp
      #briere_time1 = briere_time1, # variable temp
      #briere_value1 = briere_value1,
      #briere_time2 = briere_time2,
      #briere_value2 = briere_value2,
      #briere_time3 = briere_time3,
      #briere_value3 = briere_value3,
      #briere_time4 = briere_time4,
      #briere_value4 = briere_value4,
      #briere_time5 = briere_time5,
      #briere_value5 = briere_value5,
      seed_pop=seed_pop
    )
  )
  
  dust_system_set_state_initial(sys)
  
  t <- seq(0, 3649) # 10 years 
  
  y <- dust_system_simulate(sys, t)
  
  y <- dust_unpack_state(sys, y)
  
  
  
  ################################
  # proportion of infected birds #
  ################################
  
  
  
  df <- as.data.frame(t(y$I))
  df$t <- t
  colnames(df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")
  df_long <- pivot_longer(df, cols = c("P1", "P2", "P3", "P4", "P5"),
                          names_to = "Population",
                          values_to = "I")
  
  
  prop <- 100 * y$I / (y$S + y$E + y$I + y$R) # proportion infected
  
  df <- as.data.frame(t(prop))
  df$t <- t
  colnames(df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")
  
  df_propI_long <- pivot_longer(df,
                                cols = c(P1, P2, P3, P4, P5),
                                names_to = "Population",
                                values_to = "PropI")
  
  # save
  # Peak summary for this simulation
  peak_results[[rep]] <- df_propI_long %>%
    mutate(
      Year = floor(t / 365) + 1,
      Seed = as.integer(
        Population == paste0("P", seed_pop[Year])
      ),
      Simulation = rep
    ) %>%
    group_by(
      Population,
      Year,
      Seed,
      Simulation
    ) %>%
    summarise(
      PeakInfected = round(max(PropI), 1),
      PeakDay = ifelse(
        round(max(PropI), 1) > 0,
        (t[which.max(PropI)] %% 365) + 1,
        NA
      ),
      .groups = "drop"
    )
  
  # figure 17
  trajectory_results[[rep]] <- df_propI_long %>%
    mutate(
      Year = floor(t / 365) + 1,
      Seed = as.integer(
        Population == paste0("P", seed_pop[Year])
      ),
      Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded"),
      Simulation = rep,
      Scenario = scenario_name
    )
  
  # Combine all 50 simulations
  peak_summary <- dplyr::bind_rows(peak_results)
  #dem_trajectory <- bind_rows(trajectory_results)
}
```
## distance-dependent model
```{r distance depndent model, echo=FALSE}
sir <- odin({
  
  # Equations for transitions between compartments by group
  
  deriv(S[]) <- (gamma[i]*N[i]) - (lambda[i]*S[i]) - (mu*S[i]) - seed[i]
  
  deriv(E[]) <- (lambda[i]*S[i]) - (theta*E[i]) - (mu*E[i])
  
  deriv(I[]) <- (theta*E[i]) - (mu*I[i]) - (sigma*I[i]) + seed[i]
  
  deriv (R[]) <- (sigma*I[i]) - (mu*R[i])
  
  deriv(incidence[]) <- theta * E[i]
  
  # beta
  
  beta0 <- parameter(0.5)
  betaT <- briere * beta0 
  h0 <- parameter(0.5) # the hazard in the immediate vicinity of the host
  r0 <- parameter(50) # the distance at which the hazard is half of the maximal hazard (assuming max distance 180)
  alpha <- parameter(2) # decay parameter which determines shape of kernel
  beta[,] <- (h0 / (1 + (distance[i,j] / r0)^alpha))
  new_beta[,] <- if (i == j) betaT else beta[i, j]

  
  lambda[] <- (
    new_beta[i,1] * I[1]/N[i] +
    new_beta[i,2] * I[2]/N[i] +
    new_beta[i,3] * I[3]/N[i] +
    new_beta[i,4] * I[4]/N[i] +
    new_beta[i,5] * I[5]/N[i]
  )
  
  
  
  # initial conditions
  
  N[] <- S[i] + E[i] + I[i] + R[i]
  
  initial(S[]) <- 500 
  
  initial(E[]) <- 0
  
  initial(I[]) <- 0 
  
  initial(R[]) <- 0
  
  initial(incidence[]) <- 0
  
  # seed in july every year
  year <- floor(time/365) + 1
  seed[] <- if(
    i == seed_pop[year] &&
    time%%365 >= 182 && # July
    time%%365 < 183 
  ) 1 else 0
  
  # User defined parameters:
  
  briere <- interpolate(briere_time, briere_value, "linear")
  
  briere_time <- parameter(constant = TRUE)
  briere_value <- parameter(constant = TRUE)
  
  seed_pop <- parameter(constant = TRUE)
  
  mu <- parameter(0.000913242) # death rate
  
  theta <- parameter(0.5) # incubation period
  
  sigma <- parameter(0.25) # recovery rate
  
  distance <- parameter(constant = TRUE)

  
  
  # birth rate
  gamma[] <- if (time %% 365 >= 90 && time %% 365 <= 212) 0.0027548209 else 0 
  # 2 per bird may-july
  
  # Dimensions of arrays
  
  dim(S) <- 5
  
  dim(E) <- 5
  
  dim(I) <- 5
  
  dim(R) <- 5
  
  dim(N) <- 5
  
  dim(incidence) <- 5
  
  dim(new_beta) <- c(5, 5)
  
  dim(beta) <- c(5, 5)
  
  dim(distance) <- c(5,5)
  
  dim(briere_time, briere_value) <- parameter(rank = 1)
  
  dim(lambda) <- 5
  
  dim(seed_pop) <- 10
  
  dim(seed) <- 5
  
  dim(gamma) <- 5
})

##############
# Simulation #
##############

sys <- dust_system_create(
  sir,
  list(
    beta = beta,
    briere_time = briere_time,
    briere_value = briere_value,
    seed_pop=seed_pop,
    distance=distance
    
  )
)

dust_system_set_state_initial(sys)

t <- seq(0, 3649) # 10 years

y <- dust_system_simulate(sys, t)

y <- dust_unpack_state(sys, y)
```

## prevalence plot

```{r prevalence plot, echo=FALSE}

pdf_plotI <- df_propI_long[df_propI_long$t >= (365 * 4), ]  # plot last 6 years

## 6 years ##


month_breaks <- c(0,31,59,90,120,151,181,212,243,273,304,334)

ggplot(data = subset(df_plotI), aes(x = t %% 365, y = PropI, color = Population)) +
  geom_line() +
  facet_wrap(~(floor(t/365)+1), ncol = 1) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = c("Jan","Feb","Mar","Apr","May","Jun",
               "Jul","Aug","Sep","Oct","Nov","Dec"),
    expand = c(0, 0)
  ) +
  labs(
    x = NULL,
    y = "Prevalence (%)"
  ) +
  theme_minimal()+
  theme(
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )
```
## multi-panel plot

```{r multi panel plot, echo=FALSE}
# Proportion of susceptible birds

S_df <- as.data.frame(t(y$S))
S_df$t <- t

colnames(S_df)[1:5] <- c(
  "P1",
  "P2",
  "P3",
  "P4",
  "P5"
)

S_long <- pivot_longer(
  S_df,
  cols = c("P1","P2","P3","P4","P5"),
  names_to = "Population",
  values_to = "S"
)

propS <- 100 * y$S / (y$S + y$E + y$I + y$R)

df <- as.data.frame(t(propS))
df$t <- t

colnames(df)[1:5] <- c("P1", "P2", "P3", "P4", "P5")

df$Year <- floor(df$t / 365) + 1
df$Day <- df$t %% 365

df_propS_long <- pivot_longer(
  df,
  cols = c(P1, P2, P3, P4, P5),
  names_to = "Population",
  values_to = "PropS"
)

# Total birds

N_mat <- y$S + y$E + y$I + y$R

N_df <- as.data.frame(t(N_mat))
N_df$t <- t

colnames(N_df)[1:5] <- c(
  "P1",
  "P2",
  "P3",
  "P4",
  "P5"
)

N_long <- pivot_longer(
  N_df,
  cols = c("P1",
           "P2",
           "P3",
           "P4",
           "P5"),
  names_to = "Population",
  values_to = "N"
)

#########
# plots #
#########

# panel a

seed_data <- expand.grid(
  Year = 1:10,
  Population = 1:5
)

seed_data$Seed <- as.integer(
  seed_data$Population == seed_pop[seed_data$Year]
)

seed_panel <- ggplot(seed_data, aes(x = Year, y = Seed, colour = factor(Year))) +
  geom_point(size = 1.5) +
  facet_wrap(~ Population, nrow = 1) +
  scale_colour_manual(
    values = colorRampPalette(
      c("#BDD7E7", "#08519C"))(10),
    name="Year"
  )+
  scale_x_continuous(
    breaks = 1:10,
    expand = c(0.05, 0.05)
  ) +
  scale_y_continuous(
    breaks = c(0, 1),
    labels = c("No", "Yes"),
    limits = c(-0.1, 1.1),
    expand = c(0, 0)
  ) +
  labs(
    x = "Year",
    y = "Seeded"
  ) +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.spacing = unit(0.8, "lines"),
    legend.position = "none"
  )


# Panel b

df_propI_long <- df_propI_long %>%
  mutate(
    Briere = rep(briere_value, each = 5),
    Year = floor(t / 365) + 1,
    Day = t %% 365
  ) %>%
  filter(Briere > 0)

month_breaks <- c(0, 31, 61, 92, 122, 153, 183, 214, 245, 275, 306, 336)

month_labels <- c(
  "J", "F", "M", "A", "M", "J",
  "J", "A", "S", "O", "N", "D"
)

prevalence_panel <- ggplot(
  df_propI_long,
  aes(
    x = Day,
    y = PropI,
    colour = factor(Year),
    group = Year
  )
) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~ Population, nrow = 1) +
  scale_colour_manual(
    values = colorRampPalette(
      c("#BDD7E7", "#08519C")
    )(10)
  ) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels
  ) +
  labs(
    x = "Month",
    y = "Infection prevalence (%)",
    colour = "Year"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank()
  )

# Panel c

susceptible_panel <- ggplot(
  df_propS_long,
  aes(
    x = Day,
    y = PropS,
    colour = factor(Year),
    group = Year
  )
) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~ Population, nrow = 1) +
  scale_colour_manual(
    values = colorRampPalette(
      c("#BDD7E7", "#08519C")
    )(10)
  ) +
  scale_x_continuous(
    breaks = month_breaks,
    labels = month_labels
  ) +
  labs(
    x = "Month",
    y = "Proportion susceptible (%)",
    colour = "Year"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    legend.position = "none"
  )

# Panel d

Incidence <- as.data.frame(t(y$incidence))

Incidence$t <- t

colnames(Incidence)[1:5] <- c("P1", "P2", "P3", "P4", "P5")

Incidence_long <- Incidence %>%
  pivot_longer(
    cols = c("P1", "P2", "P3", "P4", "P5"),
    names_to = "Population",
    values_to = "Incidence"
  )

# Add year and Brière
Incidence_long$Year <- floor(Incidence_long$t / 365) + 1
Incidence_long$Briere <- rep(briere_value, each = 5)

S_long$Year <- floor(S_long$t / 365) + 1
S_long$Briere <- rep(briere_value, each = 5)

N_long$Year <- floor(N_long$t / 365) + 1
N_long$Briere <- rep(briere_value, each = 5)


# Seasonal incidence

AR <- Incidence_long %>%
  filter(Briere > 0) %>%
  group_by(Population, Year) %>%
  summarise(
    Incidence = last(Incidence) - first(Incidence),
    .groups = "drop"
  )

# Susceptible at start of season

S_start <- S_long %>%
  filter(Briere > 0) %>%
  group_by(Population, Year) %>%
  summarise(
    S_start = first(S),
    .groups = "drop"
  )

# Cumulative births during the season

Births <- N_long %>%
  filter(
    Briere > 0,
    (t %% 365) >= 90,
    (t %% 365) <= 212
  ) %>%
  mutate(
    Births_at_time = 0.0027322404 * N
  ) %>%
  group_by(Population, Year) %>%
  summarise(
    Births = sum(Births_at_time),
    .groups = "drop"
  )


# Combine incidence, S at start, and births

AR <- AR %>%
  left_join(
    S_start,
    by = c("Population", "Year")
  ) %>%
  left_join(
    Births,
    by = c("Population", "Year")
  )

# denominator = susceptible at start + cumulative births during season

AR <- AR %>%
  mutate(
    AttackRate = 100 * Incidence / (S_start + Births)
  )

# Plot

attack_rate <- ggplot(
  AR,
  aes(x = Year, y = AttackRate, fill = factor(Year))
) +
  geom_col() +
  facet_wrap(~ Population, nrow = 1) +
  scale_fill_manual(
    values = colorRampPalette(
      c("#BDD7E7", "#08519C")
    )(10),
    name = "Year"
  ) +
  scale_x_continuous(
    breaks = 1:10
  ) +
  labs(
    x = "Year",
    y = "Seasonal attack rate (%)"
  ) +
  theme_minimal()+
  theme(
    legend.position = "none")


# Panel

(seed_panel)/
  (prevalence_panel)/
  (susceptible_panel)/
  (attack_rate)+
  plot_layout(heights = c(1, 2, 2, 2))+
  plot_annotation(tag_levels = "a")

```
## plot of mean peak prevalence and day

```{r mean prevalence and day, echo=FALSE}
# prevalence

peak_plot_data <- peak_summary %>%
  filter(PeakDay != 182) %>%
  group_by(Year, Seed) %>%
  summarise(
    MeanPeak = mean(PeakInfected, na.rm = TRUE),
    MinPeak = min(PeakInfected, na.rm = TRUE),
    MaxPeak = max(PeakInfected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded")
  )

peak_plot <- ggplot(peak_plot_data,
                    aes(x = Year,
                        y = MeanPeak,
                        fill = Seeded)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_errorbar(
    aes(ymin = MinPeak,
        ymax = MaxPeak),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  scale_fill_manual(
    values = c(
      "Seeded" = "#08519C",
      "Non-seeded" = "grey85"
    )
  ) +
  labs(
    x = "Year",
    y = "Peak prevalence (%)",
    fill = NULL
  ) +
  scale_x_continuous(breaks = 1:10, labels = 1:10)+
  theme_minimal()+ 
  theme(
    legend.position = "bottom")

# peak day with dodge

peak_day_plot_data <- peak_summary %>%
  filter(PeakDay > 184) %>%
  group_by(Year, Seed) %>%
  summarise(
    MeanPeakDay = mean(PeakDay, na.rm = TRUE),
    MinPeakDay = min(PeakDay, na.rm = TRUE),
    MaxPeakDay = max(PeakDay, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded")
  )

peak_day_plot <- ggplot(
  peak_day_plot_data,
  aes(
    x = Year,
    y = MeanPeakDay,
    shape = Seeded,
    group = Seeded
  )
) +
  geom_errorbar(
    aes(
      ymin = MinPeakDay,
      ymax = MaxPeakDay
    ),
    width = 0.15,
    position = position_dodge(width = 0.25)
  ) +
  geom_point(
    size = 1.5,
    position = position_dodge(width = 0.25)
  ) +
  labs(
    x = "Year",
    y = "Peak day",
    shape = NULL
  ) +
  scale_x_continuous(
    breaks = 1:10,
    labels = 1:10
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )


# Panel
(peak_plot / peak_day_plot)

overall_mean_peak <- peak_summary %>%
  filter(
    PeakDay != 182,
    Year != 1
  ) %>%
  group_by(Seed) %>%
  summarise(
    MeanPeak = mean(PeakInfected, na.rm = TRUE),
    MinPeak = min(PeakInfected, na.rm = TRUE),
    MaxPeak = max(PeakInfected, na.rm = TRUE),
    n = sum(!is.na(PeakInfected)),
    .groups = "drop"
  ) %>%
  mutate(
    Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded")
  )

overall_mean_peak_day <- peak_summary %>%
  filter(
    PeakDay > 184,
    Year != 1
  ) %>%
  group_by(Seed) %>%
  summarise(
    MeanPeakDay = mean(PeakDay, na.rm = TRUE),
    MinPeakDay = min(PeakDay, na.rm = TRUE),
    MaxPeakDay = max(PeakDay, na.rm = TRUE),
    n = sum(!is.na(PeakDay)),
    .groups = "drop"
  ) %>%
  mutate(
    Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded")
  )
```
## figure 17
```{r figure 17, echo=FALSE}
# prevalence

peak_plot_data <- peak_summary %>%
  filter(PeakDay != 182) %>%
  group_by(Year, Seed) %>%
  summarise(
    MeanPeak = mean(PeakInfected, na.rm = TRUE),
    MinPeak = min(PeakInfected, na.rm = TRUE),
    MaxPeak = max(PeakInfected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded")
  )

peak_plot <- ggplot(peak_plot_data,
                    aes(x = Year,
                        y = MeanPeak,
                        fill = Seeded)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_errorbar(
    aes(ymin = MinPeak,
        ymax = MaxPeak),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  scale_fill_manual(
    values = c(
      "Seeded" = "#08519C",
      "Non-seeded" = "grey85"
    )
  ) +
  labs(
    x = "Year",
    y = "Peak prevalence (%)",
    fill = NULL
  ) +
  scale_x_continuous(breaks = 1:10, labels = 1:10)+
  theme_minimal()+ 
  theme(
    legend.position = "bottom")

# peak day with dodge

peak_day_plot_data <- peak_summary %>%
  filter(PeakDay > 184) %>%
  group_by(Year, Seed) %>%
  summarise(
    MeanPeakDay = mean(PeakDay, na.rm = TRUE),
    MinPeakDay = min(PeakDay, na.rm = TRUE),
    MaxPeakDay = max(PeakDay, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded")
  )

peak_day_plot <- ggplot(
  peak_day_plot_data,
  aes(
    x = Year,
    y = MeanPeakDay,
    shape = Seeded,
    group = Seeded
  )
) +
  geom_errorbar(
    aes(
      ymin = MinPeakDay,
      ymax = MaxPeakDay
    ),
    width = 0.15,
    position = position_dodge(width = 0.25)
  ) +
  geom_point(
    size = 1.5,
    position = position_dodge(width = 0.25)
  ) +
  labs(
    x = "Year",
    y = "Peak day",
    shape = NULL
  ) +
  scale_x_continuous(
    breaks = 1:10,
    labels = 1:10
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )


# Panel
(peak_plot / peak_day_plot)

overall_mean_peak <- peak_summary %>%
  filter(
    PeakDay != 182,
    Year != 1
  ) %>%
  group_by(Seed) %>%
  summarise(
    MeanPeak = mean(PeakInfected, na.rm = TRUE),
    MinPeak = min(PeakInfected, na.rm = TRUE),
    MaxPeak = max(PeakInfected, na.rm = TRUE),
    n = sum(!is.na(PeakInfected)),
    .groups = "drop"
  ) %>%
  mutate(
    Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded")
  )

overall_mean_peak_day <- peak_summary %>%
  filter(
    PeakDay > 184,
    Year != 1
  ) %>%
  group_by(Seed) %>%
  summarise(
    MeanPeakDay = mean(PeakDay, na.rm = TRUE),
    MinPeakDay = min(PeakDay, na.rm = TRUE),
    MaxPeakDay = max(PeakDay, na.rm = TRUE),
    n = sum(!is.na(PeakDay)),
    .groups = "drop"
  ) %>%
  mutate(
    Seeded = ifelse(Seed == 1, "Seeded", "Non-seeded")
  )

baseline_trajectory <- baseline_trajectory %>%
  mutate(Scenario = "Baseline")

trajectory_summary <- trajectory_summary %>%
  mutate(Scenario = "Temperature")

all_trajectories <- bind_rows(
  baseline_trajectory,
  trajectory_summary
)


trajectory_plot_data <- all_trajectories %>%
  filter(Year >= 2) %>%
  group_by(Simulation, Scenario, Year, t, Seeded) %>%
  summarise(
    MeanPropI = mean(PropI, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Scenario, Year, t, Seeded) %>%
  summarise(
    Mean = mean(MeanPropI, na.rm = TRUE),
    Lower = quantile(MeanPropI, 0.025, na.rm = TRUE),
    Upper = quantile(MeanPropI, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

year_breaks <- c(
  365, 730, 1095, 1460, 1825,
  2190, 2555, 2920, 3285
)

year_labels <- c(
  "2", "3", "4", "5", "6",
  "7", "8", "9", "10")


temp_dynamics <- ggplot(
  trajectory_plot_data,
  aes(x = t, y = Mean, colour = Scenario)
) +
  geom_ribbon(
    aes(
      ymin = Lower,
      ymax = Upper,
      fill = Scenario
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Seeded, ncol = 1) +
  scale_x_continuous(
    breaks = year_breaks,
    labels = year_labels
  ) +
  labs(
    x = "Time (years)",
    y = "Mean prevalence (%)",
    colour = NULL,
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside", legend.position.inside = c(0.85, 0.95)
  )



# population size trajectory

popsize_trajectory <- popsize_trajectory %>%
  mutate(Scenario = "Population size")

all_trajectories <- bind_rows(
  baseline_trajectory,
  popsize_trajectory
)


trajectory_plot_data <- all_trajectories %>%
  filter(Year >= 2) %>%
  group_by(Simulation, Scenario, Year, t, Seeded) %>%
  summarise(
    MeanPropI = mean(PropI, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Scenario, Year, t, Seeded) %>%
  summarise(
    Mean = mean(MeanPropI, na.rm = TRUE),
    Lower = quantile(MeanPropI, 0.025, na.rm = TRUE),
    Upper = quantile(MeanPropI, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

year_breaks <- c(
  365, 730, 1095, 1460, 1825,
  2190, 2555, 2920, 3285
)

year_labels <- c(
  "2", "3", "4", "5", "6",
  "7", "8", "9", "10"
)


pop_dynamics <- ggplot(
  trajectory_plot_data,
  aes(x = t, y = Mean, colour = Scenario)
) +
  geom_ribbon(
    aes(
      ymin = Lower,
      ymax = Upper,
      fill = Scenario
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Seeded, ncol = 1) +
  scale_x_continuous(
    breaks = year_breaks,
    labels = year_labels
  ) +
  labs(
    x = "Time (years)",
    y = "Mean prevalence (%)",
    colour = NULL,
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside", legend.position.inside = c(0.85, 0.95)
    )



# mean infection dynamics over time (scenarios 8-10 compared to baseline)

dem_trajectory <- dem_trajectory %>%
  mutate(Scenario = "Demographic traits")

all_trajectories <- bind_rows(
  baseline_trajectory,
  dem_trajectory
)


trajectory_plot_data <- all_trajectories %>%
  filter(Year >= 2) %>%
  group_by(Simulation, Scenario, Year, t, Seeded) %>%
  summarise(
    MeanPropI = mean(PropI, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Scenario, Year, t, Seeded) %>%
  summarise(
    Mean = mean(MeanPropI, na.rm = TRUE),
    Lower = quantile(MeanPropI, 0.025, na.rm = TRUE),
    Upper = quantile(MeanPropI, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

year_breaks <- c(
  365, 730, 1095, 1460, 1825,
  2190, 2555, 2920, 3285
)

year_labels <- c(
  "2", "3", "4", "5", "6",
  "7", "8", "9", "10"
)

# edit code for appropriate scenario

dem_dynamics <- ggplot(
  trajectory_plot_data,
  aes(x = t, y = Mean, colour = Scenario)
) +
  geom_ribbon(
    aes(
      ymin = Lower,
      ymax = Upper,
      fill = Scenario
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Seeded, ncol = 1) +
  scale_x_continuous(
    breaks = year_breaks,
    labels = year_labels
  ) +
  labs(
    x = "Time (years)",
    y = "Mean prevalence (%)",
    colour = NULL,
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside", legend.position.inside = c(0.85, 0.95)
  )

# panel
(pop_dynamics/
    dem_dynamics/
    temp_dynamics)+
  plot_annotation(tag_levels = "a")


# colour-coded attack rates compared to baseline


## attack rate saved from multi-panel figures

# TEMPERATURE

AR_temp <- AR_temp %>%
  mutate(Scenario = "Temperature")

AR_baseline <- AR_baseline %>%
  mutate(Scenario = "Baseline")


# Bind baseline and temperature data
AR_base_temp <- bind_rows(
  AR_baseline,
  AR_temp
)


# Calculate whether AR is higher/lower than baseline
AR_plot <- AR_base_temp %>%
  group_by(Population, Year) %>%
  mutate(
    BaselineAR = AttackRate[Scenario == "Baseline"],
    Direction = case_when(
      Scenario == "Baseline" ~ "Baseline",
      AttackRate > BaselineAR ~ "Higher",
      AttackRate < BaselineAR ~ "Lower",
      TRUE ~ "Same"
    )
  ) %>%
  ungroup()


AR_temp_plot <- AR_plot %>%
  filter(
    Scenario == "Temperature"
  )


AR_temp_plot <- AR_temp %>%
  left_join(
    AR_baseline %>%
      select(Population, Year, BaselineAR = AttackRate),
    by = c("Population", "Year")
  ) %>%
  mutate(
    Difference = AttackRate - BaselineAR
  )

# plot
AR_temp <- ggplot(
  AR_temp_plot,
  aes(
    x = Year,
    y = AttackRate,
    fill = Difference
  )
) +
  geom_col() +
  facet_wrap(
    ~ Population,
    nrow = 1
  ) +
  scale_fill_gradient2(
    low = "#4575B4",
    mid = "grey85",
    high = "#D73027",
    midpoint = 0,
    limits = c(-20, 20),
    name = "Difference"
  ) +
  scale_x_continuous(
    breaks = 1:10
  ) +
  labs(
    x = "Year",
    y = "Seasonal attack rate (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )


# POPULATION SIZE


AR_pop <- AR_pop %>%
  mutate(Scenario = "Population")

AR_baseline <- AR_baseline %>%
  mutate(Scenario = "Baseline")


# Bind baseline and population size data
AR_base_pop <- bind_rows(
  AR_baseline,
  AR_pop
)


# AR higher/lower than baseline
AR_plot <- AR_base_pop %>%
  group_by(Population, Year) %>%
  mutate(
    BaselineAR = AttackRate[Scenario == "Baseline"],
    Direction = case_when(
      Scenario == "Baseline" ~ "Baseline",
      AttackRate > BaselineAR ~ "Higher",
      AttackRate < BaselineAR ~ "Lower",
      TRUE ~ "Same"
    )
  ) %>%
  ungroup()

AR_pop_plot <- AR_plot %>%
  filter(
    Scenario == "Population"
  )

AR_pop_plot <- AR_pop %>%
  left_join(
    AR_baseline %>%
      select(Population, Year, BaselineAR = AttackRate),
    by = c("Population", "Year")
  ) %>%
  mutate(
    Difference = AttackRate - BaselineAR
  )


AR_pop <-ggplot(
  AR_pop_plot,
  aes(
    x = Year,
    y = AttackRate,
    fill = Difference
  )
) +
  geom_col() +
  facet_wrap(
    ~ Population,
    nrow = 1
  ) +
  scale_fill_gradient2(
    low = "#4575B4",
    mid = "grey85",
    high = "#D73027",
    midpoint = 0,
    limits = c(-20, 20),
    name = "Difference"
  ) +
  scale_x_continuous(
    breaks = 1:10
  ) +
  labs(
    x = "Year",
    y = "Seasonal attack rate (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )

# DEMOGRAPHY

AR_dem <- AR_dem %>%
  mutate(Scenario = "Demographic traits")

AR_baseline <- AR_baseline %>%
  mutate(Scenario = "Baseline")


# Bind baseline and demography data
AR_base_dem <- bind_rows(
  AR_baseline,
  AR_dem
)


# AR higher/lower than baseline
AR_plot <- AR_base_dem %>%
  group_by(Population, Year) %>%
  mutate(
    BaselineAR = AttackRate[Scenario == "Baseline"],
    Direction = case_when(
      Scenario == "Baseline" ~ "Baseline",
      AttackRate > BaselineAR ~ "Higher",
      AttackRate < BaselineAR ~ "Lower",
      TRUE ~ "Same"
    )
  ) %>%
  ungroup()

AR_dem_plot <- AR_plot %>%
  filter(
    Scenario == "Demographic traits"
  )

AR_dem_plot <- AR_dem %>%
  left_join(
    AR_baseline %>%
      select(Population, Year, BaselineAR = AttackRate),
    by = c("Population", "Year")
  ) %>%
  mutate(
    Difference = AttackRate - BaselineAR
  )


AR_dem <- ggplot(
  AR_dem_plot,
  aes(
    x = Year,
    y = AttackRate,
    fill = Difference
  )
) +
  geom_col() +
  facet_wrap(
    ~ Population,
    nrow = 1
  ) +
  scale_fill_gradient2(
    low = "#4575B4",
    mid = "grey85",
    high = "#D73027",
    midpoint = 0,
    limits = c(-20, 20),
    name = "Difference"
  ) +
  scale_x_continuous(
    breaks = 1:10
  ) +
  labs(
    x = "Year",
    y = "Seasonal attack rate (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )

# panel
(AR_pop/
    AR_dem/
    AR_temp)+
  plot_annotation(tag_levels = "a")

(pop_dynamics/
    dem_dynamics/
    temp_dynamics)+
  plot_annotation(tag_levels = "a")

(
  (AR_pop | pop_dynamics) /
    (AR_dem | dem_dynamics) /
    (AR_temp | temp_dynamics)
) +
  plot_annotation(tag_levels = "a")


```
