---
title: "Temperature model with third population"
output: html_document
date: "2026-06-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Equations

$$
\frac{\mathrm{d}S_i}{\mathrm{d}t} = (\gamma N_i) - ((m_{i,1} \frac {I_1}{N_1} + m_{i,2} \frac {I_2}{N_2} + m_{i,3} \frac {I_3}{N_3})S_i) - (\mu S_i)
$$ $$
\frac{\mathrm{d}E_i}{\mathrm{d}t} = ((m_{i,1} \frac {I_1}{N_1} + m_{i,2} \frac {I_2}{N_2} + m_{i,3} \frac {I_3}{N_3})S_i) - (\theta E_i) - (\mu E_i)
$$ $$
\frac{\mathrm{d}I_i}{\mathrm{d}t} = (\theta E_i) - (\mu I_i) - (\sigma I_i)
$$ $$
\frac{\mathrm{d}R_i}{\mathrm{d}t} = (\sigma I_i) - (\mu R_i)
$$

## Matrix

$$
\begin{aligned}
m &=
\begin{bmatrix}
0.9 & 0.1 & 0.1 \\
0.1 & 0.6 & 0.1 \\
0.1 & 0.1 & 0.3
\end{bmatrix}
\qquad
m &=
\begin{bmatrix}
\beta_{ii} & \beta_{ij} & \beta_{ik}\\
\beta_{ji} & \beta_{jj} & \beta_{jk}\\
\beta_{ki} & \beta_{kj} & \beta_{kk}
\end{bmatrix}
\end{aligned}
$$

This is meant to simulate 3 populations with minimal transmission between them, and different transmission rates in each population

```{r matrix}
library(odin2)
library(dust2)
library(monty)
library(ggplot2)
library(tidyr)

m <- matrix(c(0.9,0.1,0.1,0.1,0.6,0.1,0.1,0.1,0.3), 3, 3, byrow=TRUE)
```

## SIR model

I used exp(alpha\*temp) to include temperature as a basic example, to follow the temperature dependence set by Marini et al the mosquito population is needed

I also had to decrease the avian fertility rate from 0.05 to 0.005 for the code to run, I'm not sure why

```{r SIR model, echo=TRUE}
sir <- odin({
  
  # Equations for transitions between compartments by group
  
  deriv(S[]) <- (gamma*N[i]) - (beta_temp*(m[i,1] * I[1]/N[1] + m[i,2] * I[2]/N[2] + m[i,3] * I[3]/N[3])*S[i]) - (mu*S[i])
  
  deriv(E[]) <- (beta_temp*(m[i,1] * I[1]/N[1] + m[i,2] * I[2]/N[2] + m[i,3] * I[3]/N[3])*S[i]) - (theta*E[i]) - (mu*E[i])
  
  deriv(I[]) <- (theta*E[i]) - (mu*I[i]) - (sigma*I[i])
  
  deriv (R[]) <- (sigma*I[i]) - (mu*R[i])
  
  # initial conditions
  
  N[] <- S[i] + E[i] + I[i] + R[i] # Total number of birds in each population
  
  initial(S[1]) <- 500 - I0
  initial(S[2]) <- 500 
  initial(S[3]) <- 500 
  
  initial(E[]) <- 0
  
  initial(I[1]) <- I0
  initial(I[2]) <- 0
  initial(I[3]) <- 0
  
  initial(R[]) <- 0
  
  # User defined parameters:
  
  m <- parameter()
  
  I0 <- parameter(1) # starts with 1 infected bird
  
  alpha <- 0.05 # temperature 
  
  temp <- interpolate(temp_time, temp_value, "spline")
  
  temp_time <- parameter(constant = TRUE)
  
  temp_value <- parameter(constant = TRUE)
  
  beta_temp <- exp(alpha*temp)
  
  mu <- parameter(0.0015) 
  
  theta <- parameter(0.5)
  
  sigma <- parameter(0.57)
  
  gamma <- if (time %% 365 >= 184) 0.005 else 0 # should be 0.05, Jan-July (year starts in July)
  
  # Dimensions of arrays
  
  dim(S) <- 3
  
  dim(E) <- 3
  
  dim(I) <- 3
  
  dim(R) <- 3
  
  dim(N) <- 3
  
  dim(m) <- c(3, 3)
  
  dim(temp_time, temp_value) <- parameter(rank = 1)
  
})
```

## Temperature series

I created the temperature series to start in July so that the initial infection occurs at a more realistic time. I also attempted to run the model over 3 years, but I'm not sure if this is the correct approach

```{r temperature series, echo=TRUE}
base_temp_time <- c(0, 31, 59, 90, 120, 151, 181, 212, 243, 273, 304, 334, 365) # times at which temp will be specified
base_temp_value <- c(21.8, 21.3, 17.1, 13, 8.3, 3.9, 2.7, 4, 7.1, 11.3, 15.7, 20.1, 21.8) #start in July
temp_time <- c(
  base_temp_time,
  base_temp_time[-1] + 365, # year 2
  base_temp_time[-1] + 730 # year 3
)

temp_value <- c(
  base_temp_value,
  base_temp_value[-1], # year 2
  base_temp_value[-1] # year 3
)
plot(spline(temp_time, temp_value, n = 1095), type = "l", # create smooth spline curve (estimates beta between known time points)
     xlab = "Time", ylab = "temp")
points(temp_time, temp_value, pch = 12)
```

## Plot of infected birds

From this output I think the model is not working correctly (a year is skipped) but I'm not sure why

```{r infected birds, echo=TRUE}
sys <- dust_system_create(
  sir,
  list(
    m = m,
    temp_time = temp_time,
    temp_value = temp_value
  )
)

dust_system_set_state_initial(sys)

t <- seq(0, 1095) # 3 years

y <- dust_system_simulate(sys, t)

y <- dust_unpack_state(sys, y)

R_df <- as.data.frame(t(y$R))
R_df$t <- t
colnames(R_df)[1:3] <- c("population 1", "population 2", "population 3")

df <- as.data.frame(t(y$I))
df$t <- t
colnames(df)[1:3] <- c("population 1", "population 2", "population 3")
df_long <- pivot_longer(df, cols = c("population 1", "population 2", "population 3"),
                        names_to = "group",
                        values_to = "I")
ggplot(df_long, aes(x = t, y = I, color = group)) +
  geom_line() +
  labs(title = "P1 beta=0.9, P2 beta=0.6, P3 beta = 0.3", x = "Time (days)", y = "Infected birds") +
  theme_minimal()
```

## Plot of recovered birds

```{r recovered birds, echo=TRUE}
R_long <- pivot_longer(
  R_df,
  cols = c("population 1", "population 2", "population 3"),
  names_to = "group",
  values_to = "R"
)

ggplot(R_long, aes(x = t, y = R, color = group)) +
  geom_line() +
  labs(title = "Recovered birds over time",
       x = "Time (days)",
       y = "Recovered (R)") +
  theme_minimal()
```

## Plot of total birds

```{r total birds, echo=TRUE}

N_mat <- y$S + y$E + y$I + y$R

N_df <- as.data.frame(t(N_mat))
N_df$t <- t

colnames(N_df)[1:3] <- c(
  "population 1",
  "population 2",
  "population 3"
)

N_long <- pivot_longer(
  N_df,
  cols = c("population 1",
           "population 2",
           "population 3"),
  names_to = "group",
  values_to = "N"
)

ggplot(N_long, aes(x = t, y = N, color = group)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Total population size over time",
    x = "Time (days)",
    y = "Number of birds"
  ) +
  theme_minimal()

```

## Plot of susceptible birds

```{r susceptible birds, echo=TRUE}

S_df <- as.data.frame(t(y$S))
S_df$t <- t

colnames(S_df)[1:3] <- c(
  "population 1",
  "population 2",
  "population 3"
)

S_long <- pivot_longer(
  S_df,
  cols = c("population 1","population 2","population 3"),
  names_to = "group",
  values_to = "S"
)

ggplot(S_long,
       aes(t, S, colour = group)) +
  geom_line()

```
