---
title: "updated temperature function + variable fertility rate"
output: html_document
date: "2026-06-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Equations

$$
\frac{\mathrm{d}S_i}{\mathrm{d}t} = (\gamma N_i) - (briere(m_{i,1} \frac {I_1}{N_1} + m_{i,2} \frac {I_2}{N_2} + m_{i,3} \frac {I_3}{N_3})S_i) - (\mu S_i)
$$ $$
\frac{\mathrm{d}E_i}{\mathrm{d}t} = (briere(m_{i,1} \frac {I_1}{N_1} + m_{i,2} \frac {I_2}{N_2} + m_{i,3} \frac {I_3}{N_3})S_i) - (\theta E_i) - (\mu E_i)
$$ $$
\frac{\mathrm{d}I_i}{\mathrm{d}t} = (\theta E_i) - (\mu I_i) - (\sigma I_i)
$$ $$
\frac{\mathrm{d}R_i}{\mathrm{d}t} = (\sigma I_i) - (\mu R_i)
$$

## Matrix

$$
\begin{aligned}
m &=
\begin{bmatrix}
0.9 & 0.1 & 0.1 \\
0.1 & 0.6 & 0.1 \\
0.1 & 0.1 & 0.3
\end{bmatrix}
\qquad
m &=
\begin{bmatrix}
\beta_{ii} & \beta_{ij} & \beta_{ik}\\
\beta_{ji} & \beta_{jj} & \beta_{jk}\\
\beta_{ki} & \beta_{kj} & \beta_{kk}
\end{bmatrix}
\end{aligned}
$$

This is meant to simulate 3 populations with minimal transmission between them, and different transmission rates in each population

```{r matrix}
library(odin2)
library(dust2)
library(monty)
library(ggplot2)
library(tidyr)

m <- matrix(c(0.9,0.1,0.1,0.1,0.6,0.1,0.1,0.1,0.3), 3, 3, byrow=TRUE)
```

## SIR model

I used a briere temperature function and made fertility rate align with death rate.


```{r SIR model, echo=TRUE}
sir <- odin({
  
  # Equations for transitions between compartments by group
  
  deriv(S[]) <- (gamma*N[i]) - (briere*(m[i,1] * I[1]/N[1] + m[i,2] * I[2]/N[2] + m[i,3] * I[3]/N[3])*S[i]) - (mu*S[i])
  
  deriv(E[]) <- (briere*(m[i,1] * I[1]/N[1] + m[i,2] * I[2]/N[2] + m[i,3] * I[3]/N[3])*S[i]) - (theta*E[i]) - (mu*E[i])
  
  deriv(I[]) <- (theta*E[i]) - (mu*I[i]) - (sigma*I[i])
  
  deriv (R[]) <- (sigma*I[i]) - (mu*R[i])
  
  m <- parameter()
  
  
  # initial conditions
  
  N[] <- S[i] + E[i] + I[i] + R[i]
  
  initial(S[1]) <- 500 - I0
  initial(S[2]) <- 500 
  initial(S[3]) <- 500 
  
  initial(E[]) <- 0
  
  initial(I[1]) <- I0
  initial(I[2]) <- 0
  initial(I[3]) <- 0
  
  initial(R[]) <- 0
  
  # User defined parameters:
  
  briere <- interpolate(briere_time, briere_value, "linear")
  
  briere_time <- parameter(constant = TRUE)
  briere_value <- parameter(constant = TRUE)
  
  
  I0 <- parameter(1)
  
  mu <- parameter(0.0015) 
  
  theta <- parameter(0.5)
  
  sigma <- parameter(0.57)
  
  gamma <- if (time %% 244 >= 61 && time %% 244 <= 153) 0.0025 else 0 #mu 62% increase (92/245)
  #may june july (3 months, 92 days)
  
  # Dimensions of arrays
  
  dim(S) <- 3
  
  dim(E) <- 3
  
  dim(I) <- 3
  
  dim(R) <- 3
  
  dim(N) <- 3
  
  
  dim(m) <- c(3, 3)
  
  dim(briere_time, briere_value) <- parameter(rank = 1)
})
```

## Including average daily temperature

Temperatures downloaded from teams chat

```{r briere function, echo=TRUE}
daily_temperatures_MO <- read.csv("~/daily_temperatures_MO.csv") # imported dataset (.csv) and only included march-october 
daily_temp <- daily_temperatures_MO$DAILY_AVG 

BriereNorm <- function(temp, T_minB, T_maxB, mB) {
  
  briere <- (temp - T_minB) * (T_maxB - temp)^mB
  
  briere[briere < 0] <- 0
  briere[is.na(briere)] <- 0
  
  briere <- briere / max(briere, na.rm = TRUE)
  
  return(briere)
}


briere_value <- BriereNorm(
  temp = daily_temp, 
  T_minB = 6.05,
  T_maxB = 31.15,
  mB = 0.5
)

briere_time <- seq(0, 244, by=1)

temp_seq <- seq(6.05, 31.15, by = 0.1)


plot(temp_seq,
     BriereNorm(temp_seq, 6.05, 31.15, 0.5),
     type = "l",
     xlab = "Temperature (°C)",
     ylab = "Brière response")
```

## Simulation

```{r simulation}
sys <- dust_system_create(
  sir,
  list(
    m = m,
    briere_time = briere_time,
    briere_value = briere_value
  )
)

dust_system_set_state_initial(sys)

t <- seq(0, 244) 

y <- dust_system_simulate(sys, t)

y <- dust_unpack_state(sys, y)

R_df <- as.data.frame(t(y$R))
R_df$t <- t
colnames(R_df)[1:3] <- c("population 1", "population 2", "population 3")
```

## Plot of infected birds

I also used a dataset with constant temperature to check that this gave the expected output (same as previous model without temperature)

```{r infected birds, echo=TRUE}
# plot

df <- as.data.frame(t(y$I))
df$t <- t
colnames(df)[1:3] <- c("population 1", "population 2", "population 3")
df_long <- pivot_longer(df, cols = c("population 1", "population 2", "population 3"),
                        names_to = "group",
                        values_to = "I")
ggplot(df_long, aes(x = t, y = I, color = group)) +
  geom_line() +
  labs(title = "P1 beta=0.9, P2 beta=0.6, P3 beta = 0.3", x = "Time (days)", y = "Infected birds") +
  theme_minimal()
```
## Plot of recovered birds

```{r recovered birds, echo=TRUE}
R_long <- pivot_longer(
  R_df,
  cols = c("population 1", "population 2", "population 3"),
  names_to = "group",
  values_to = "R"
)

ggplot(R_long, aes(x = t, y = R, color = group)) +
  geom_line() +
  labs(title = "Recovered birds over time",
       x = "Time (days)",
       y = "Recovered (R)") +
  theme_minimal()
```

## Plot of total birds

```{r total birds, echo=TRUE}

N_mat <- y$S + y$E + y$I + y$R

N_df <- as.data.frame(t(N_mat))
N_df$t <- t

colnames(N_df)[1:3] <- c(
  "population 1",
  "population 2",
  "population 3"
)

N_long <- pivot_longer(
  N_df,
  cols = c("population 1",
           "population 2",
           "population 3"),
  names_to = "group",
  values_to = "N"
)

ggplot(N_long, aes(x = t, y = N, color = group)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Total population size over time",
    x = "Time (days)",
    y = "Number of birds"
  ) +
  theme_minimal()

```

## Plot of susceptible birds

```{r susceptible birds, echo=TRUE}

S_df <- as.data.frame(t(y$S))
S_df$t <- t

colnames(S_df)[1:3] <- c(
  "population 1",
  "population 2",
  "population 3"
)

S_long <- pivot_longer(
  S_df,
  cols = c("population 1","population 2","population 3"),
  names_to = "group",
  values_to = "S"
)

ggplot(S_long,
       aes(t, S, colour = group)) +
  geom_line()

```

